---
title: "January 6th Twitter Data Analysis"
author: "Sage Fuentes"
date: "2026-05-05"
output: pdf_document
---

```{r setup, include=FALSE}
# global knitr options
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(tidyverse)
library(dplyr)
library(arrow)
library(lubridate)
library(tidytext)
library(tidylo) # for log-odds
library(textdata)
library(stringr)
library(scales) 
library(stringr)
library(tidyr)
```

## Data Cleaning

```{r eval=FALSE, echo = TRUE}
# Get a list of all CSV file paths in your folder
files <- list.files(path = "./Data", 
                    pattern = "*.csv", 
                    full.names = TRUE)
```

```{r eval=FALSE, echo = TRUE}
# 2. Read them all into one data frame
# read_csv automatically binds them together into one tibble
combinedData <- read_csv(files, id = "file_path")
glimpse(combinedData)
```
 
```{r eval=FALSE, echo = TRUE}
dateCleanedData <- combinedData %>% 
    # Original data was PST, but is being read as UTC, force the timezone
  mutate(time_col = force_tz(Date, tzone = "America/Los_Angeles")) %>%
  # Now convert to EST
  mutate(Date = with_tz(Date, tzone="America/New_York")) %>% 
  # renaming to EST for 
  rename(EST = Date)
```

Due to the original method of scraping this data being part of a larger project in which VADER and TextBlob scores were calculated for each tweet, the text available to us has been processed for this style of text evaluation. I am reanalyzing the text later due to improvements in technology since 2021, so the original VADER and TextBlob scores are not being included at this time to improve efficiency.

```{r eval=FALSE, echo = TRUE}
cleanSubsetData <- dateCleanedData %>%
  # Want to keep just the relevant information
  select(ProcessedText,
       EST,
       Outlinks,
       ReplyCount,
       RetweetCount,
       LikeCount,
       QuoteCount) %>% 
  # want to remove duplicates that are from the same time and have the same content
  distinct(ProcessedText, EST, .keep_all = TRUE) %>%
  # Last step, cleaning the text again and making a column of hashtags for individual analysis
  mutate(
    # Extract hashtags into a list column
    Hashtags = str_extract_all(ProcessedText, "#\\w+"),
    # Create a 'clean' version for word-level analysis
    TextClean = ProcessedText %>%
      str_remove_all("http\\S+|www\\S+") %>% # Remove URLs
      str_remove_all("@\\w+") %>%           # Remove Mentions
      str_replace_all("[^[:alnum:][:space:]]", " ") %>% # Remove Emojis/Special chars
      str_to_lower()
  ) %>% 
  # So I don't mix up text, making the old ProcessedText into 
  rename(Text = ProcessedText)
```


```{r eval=FALSE, echo = TRUE}
jan6onlyData <- cleanSubsetData %>% 
  filter(between(EST, as.Date('2021-01-06'), as.Date('2021-01-07')))
```



```{r eval=FALSE, echo = TRUE}
pqFullPath <- "./Data/capitolDataset"
cleanSubsetData %>% 
  write_dataset(path = pqPath, format = "parquet")
pqJanPath <- "./Data/jan6Dataset"
jan6onlyData %>% 
  write_dataset(path = pqJanPath, format = "parquet")
```

## Analysis

```{r}
pqJanPath <- "./Data/jan6Dataset"
capitolFullDF <- open_dataset(pqJanPath)
```

Trump
holds sentiment meaning that obscures from the AFINN based sentiment analyses
registers as a 'surprise' word in the nrc analysis


```{r}
# As there are some words in the analysis that might skew results, use tweetWords with anti_join in instances where that makes the most sense
customStopWords <- tibble(
  word = c("trump"
           ),
  lexicon = "custom"
)
```


```{r}
# Tokenize
tweetWords <- capitolFullDF %>%
  collect() %>% 
  # Now run the tokenization
  unnest_tokens(word, TextClean, drop = FALSE) %>%
  # Remove stop words
  anti_join(stop_words, by = "word")

tweetOverallWordFreq <- tweetWords %>% 
  count(word, sort = TRUE) %>% 
  slice_max(n, n = 15)
```


```{r}
# AFINN Sentiment Score per Word
sentimentWords <- tweetWords %>%
  select(EST, TextClean, word) %>% 
  # removing 'trump'
  anti_join(customStopWords, by = "word") %>% 
  # Acquire sentiments
  left_join(get_sentiments("afinn"), by = "word") %>%
  # regroup
  group_by(EST, TextClean) 

sentimentSummary <- sentimentWords %>%
  # Sum up the sentiment scores as per documentation
  summarize(SentimentScore = sum(value, na.rm = TRUE), .groups = "drop")

```

```{r}
# 1. Join with sentiments and categorize
wordSentimentCategories <- sentimentWords %>%
  ungroup() %>%
  # Filter out words that didn't have an AFINN match
  filter(!is.na(value)) %>%
  mutate(category = ifelse(value > 0, "Positive", "Negative"))

# 2. Get Top 10 words per category
topWordsByCat <- wordSentimentCategories %>%
  group_by(category, word) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(category) %>%
  slice_max(n, n = 10, with_ties = FALSE) %>%
  ungroup()

# Find the n of the top positive word
top_pos_value <- topWordsByCat %>%
  filter(category == "Positive") %>%
  slice_max(n, n = 1) %>%
  pull(n) %>%
  format(big.mark = ",") # This adds the comma for readability
```

Due to one of the outliers in the positive words ("supporters" having ), I used `facet_wrap` with scales set to `free` to ensure legibility in both 

```{r}
# Comparison Bar Chart
ggplot(topWordsByCat, aes(x = reorder_within(word, n, category), y = n, fill = category)) +
  geom_col(show.legend = FALSE, alpha = 0.8) +
  
  # 1. Annotate with better positioning
  geom_text(aes(label = format(n, big.mark = ",")), 
            hjust = -0.1,            # Slightly closer to the bar
            size = 3.5,              
            fontface = "bold",       
            color = "gray20") +      
  
  # 2. Allow labels to be drawn outside the panel (CRITICAL)
  coord_flip(clip = "off") + 
  
  facet_wrap(~category, scales = "free") +
  scale_x_reordered() + 
  
  # Named Colors
  scale_fill_manual(values = c("Positive" = "seagreen4", "Negative" = "firebrick3")) +
  
  # 3. Increase expansion to 30% to accommodate large numbers/outliers
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 11, color = "gray30"),
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    # 4. Increase the right margin of the entire plot as a safety net
    plot.margin = margin(10, 50, 10, 10) 
  ) +
  labs(
    title = "Top 10 Most Prevalent Sentiment-Bearing Words",
    subtitle = "Lexical prevalence within the AFINN sentiment dictionary",
    x = NULL, 
    y = "Total Frequency of Occurrences"
  )
```


Add Trump as a stop word for this analysis, potentially others

```{r}
nrcEmotions <- tweetWords %>%
  # removing 'trump'
  anti_join(customStopWords, by = "word")%>% 
  # join
  inner_join(get_sentiments("nrc"), 
             by = "word", 
             relationship = "many-to-many") %>%
  count(word, sentiment, sort = TRUE)

# Visualizing emotion distribution
# 1. Get the list of unique emotions
emotion_list <- unique(nrcEmotions$sentiment)

# 2. Loop through and print each plot
for (emo in emotion_list) {
  
  # Filter data for the specific emotion
  plot_data <- nrcEmotions %>%
    filter(sentiment == emo) %>%
    # limit to top 10
    slice_max(n, n = 10)
  
  # Create the plot
  p <- ggplot(plot_data, aes(x = n, y = reorder(word, n))) +
    # Draw the bars
    geom_col(aes(fill = n), show.legend = FALSE) + 
    geom_text(aes(label = comma(n)), 
              hjust = -0.1,          # Moves text slightly to the right of the bar
              size = 4,              # Adjust text size
              fontface = "bold") +
    # Added a subtle gradient fill
    scale_fill_viridis_c(option = "mako", begin = 0.3, end = 0.8) +
    # Expand x-axis so the labels don't get cut off
    scale_x_continuous(expand = expansion(mult = c(0, .15)), labels = comma) +
    theme_minimal(base_size = 14) +
    labs(
      title = paste("Top Words Expressing:", str_to_title(emo)),
      subtitle = paste("Top 10 words associated with", emo, "sentiment"),
      x = "Total Occurrences",
      y = NULL
    ) +
    theme(
      # Centering the title and subtitle
      plot.title = element_text(face = "bold", hjust = 0.5, size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray30"),
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(face = "bold")
    )

  # In Rmd loops, you must explicitly print the plot
  print(p)
}
```

```{r}
hashtagFull <- capitolFullDF %>%
  select(EST, TextClean, Hashtags) %>% 
  collect() %>% 
  filter(!is.na(Hashtags)) %>%
  # was dealing with compatibility errors with Arrow, so setting this as a vector
  mutate(Hashtags = as.vector(Hashtags)) %>% 
  # spliting each hashtag into individual rows removing the commas and spaces
  separate_rows(Hashtags, sep = ",\\s*") %>% 
  # normalizing casing
  mutate(Hashtags = tolower(Hashtags)) 

hashtagCounts <- hashtagFull %>% 
  # aggregate counts and sort by frequency for later log odds analysis
  count(Hashtags, sort = TRUE)

# View the top 10
head(hashtagCounts, 15)
```

```{r}
# join the sentiments and hashtag data using EST and TextClean to join by 
hashtagSentimentJoin <- hashtagFull %>%
  inner_join(sentimentSummary, by = c("EST", "TextClean"))

# create a list of the top 15 tags
top15Tags <- hashtagSentimentJoin %>%
  count(Hashtags, sort = TRUE) %>%
  slice_max(n, n = 15) %>%
  pull(Hashtags)

plotDataHashtagSentiment <- hashtagSentimentJoin %>%
  # remove hashtags outside of the top 15
  filter(Hashtags %in% top15Tags)
```





```{r, fig.height=5, fig.width=8}
# Create a summary table for the mean/median lines and text labels
statsLabelsHashtagSentiment <- plotDataHashtagSentiment %>%
  group_by(Hashtags) %>%
  summarize(
    mu = mean(SentimentScore, na.rm = TRUE),
    md = median(SentimentScore, na.rm = TRUE)
  )

# 2. Loop through each hashtag and generate a plot
for (current_tag in top15Tags) {
  
  # Filter data for just this hashtag
  tag_data <- hashtagSentimentJoin %>%
    filter(Hashtags == current_tag)
  
  # Calculate stats for this specific tag
  stats <- tag_data %>%
    summarize(
      n = n(),
      mu = mean(SentimentScore, na.rm = TRUE),
      md = median(SentimentScore, na.rm = TRUE),
      min_val = min(SentimentScore, na.rm = TRUE),
      max_val = max(SentimentScore, na.rm = TRUE)
    )

  # building out plots
  p <- ggplot(tag_data, aes(x = SentimentScore)) +
    geom_density(fill = "lightgoldenrod", color = "black", alpha = 0.7, adjust = 2) +
    
    geom_vline(aes(xintercept = stats$mu, color = "Mean"), linetype = "dashed", linewidth = 1) +
    geom_vline(aes(xintercept = stats$md, color = "Median"), linetype = "dotted", linewidth = 1) +
    geom_vline(xintercept = 0, color = "black", alpha = 0.3, linewidth = 0.5) +
    # Styling the Legend & Colors
    scale_color_manual(name = "Metrics", values = c("Mean" = "red", "Median" = "darkgreen")) +
    theme_minimal(base_size = 12) +
    # Center the title and customize the look
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      plot.caption = element_text(hjust = 0.5, size = 10),
      panel.grid.minor = element_blank(),
      legend.position = "top"
    ) +
    labs(
      title = paste0("Sentiment Distribution for ", current_tag),
      subtitle = "Distribution of Sentiment (Smoothed Kernel Density)",
      x = "Sentiment Score",
      y = "Density",
      caption = paste0("Total Observations (n): ", stats$n, "\n",
                       "Sentiment score - ", 
                       "Mean: ", round(stats$mu, 2),
                       " | Median: ", round(stats$md, 2),
                       " | Min: ", round(stats$min_val, 2), 
                       " | Max: ", round(stats$max_val, 2))
    )
  
  # Explicitly print the plot in each iteration
  print(p)
}
```

